---
title: "Data Basics"
author: Zach del Rosario
date: 2020-05-03
output: github_document
time: 10
reading: 0
---

# Data: Basics

*Purpose*: When first studying a new dataset, there are very simple checks we
should perform first. These are those checks.

Additionally, we'll have our first look at the *pipe operator*, which will be
super useful for writing code that's readable.

*Reading*: (None)

```{r setup}
library(tidyverse)
```

```{r options, include=FALSE}
# knitr options
knitr::opts_chunk$set(echo = TRUE)
```

## First Checks
<!-- -------------------------------------------------- -->

### __q0__ Run the following chunk:

*Hint*: You can do this either by clicking the green arrow at the top-right of
the chunk, or by using the keybaord shortcut `Shift` + `Cmd/Ctrl` + `Enter`.

```{r }
head(iris)
```

This is a *dataset*; the fundamental object we'll study throughout this course.
Some nomenclature:

- The `1, 2, 3, ...` on the left enumerate the **rows** of the dataset
- The names `Sepal.Length`, `Sepal.Width`, `...` name the **columns** of the dataset
- The column `Sepal.Length` takes **numeric** values
- The column `Species` takes **string** values

### __q1__ Load the `tidyverse` and inspect the `diamonds` dataset. What do the
`cut`, `color`, and `clarity` variables mean?

*Hint*: You can run `?diamonds` to get information on a built-in dataset.

```{r load-packages, eval=FALSE}
head(diamonds)
?diamonds
```
The cut variable refers to the quality of the cut, the color refers to the diamond color, and the clarity refers to the measurement of how clear the diamond is.

### __q2__ Run `glimpse(diamonds)`; what variables does `diamonds` have?

```{r}
glimpse(diamonds)
```
The dataset has 10 variables: carat, cut, color, clarity, depth, table, price, x, y, and z.

### __q3__ Run `summary(diamonds)`; what are the common values for each of the
variables? How widely do each of the variables vary?

*Hint*: The `Median` and `Mean` are common values, while `Min` and `Max` give us
a sense of variation.

```{r q3-task}
summary(diamonds)
```
Carat: The average carat is 0.7979 with a median of 0.7 and values range from 0.2 to 5.01.
Cut: The most common cut is ideal, followed by premium and very good. The cuts vary from fair to ideal.
Color: The most common color is G, followed by E and F. The colors vary from D to J.
Clarity: The most common clarity is SI1, followed by VS2 and SI2. The clarity values vary greatly.
Depth: The average depth is 61.75 with a median of 61.8 and values range from 43 to 79.
Table: The average table is 57.46 with a median of 61.8 and values range from 43 to 79.
Price: The average price is 3933 with a median of 2401 and values range from 326 to 18823.
x: The average x is 5.731 with a median of 5.7 and values range from 0 to 10.74.
y: The average y is 5.735 with a median of 5.71 and values range from 0 to 58.9.
z: The average z is 3.539 with a median of 3.53 and values range from 0 to 31.8.

**Observations**:

- (Write your observations here!)
There does not appear to be many diamonds with cuts below "very good" with most diamonds having a G color. The size of diamond carats seem to be fairly evenly spread out in b/w the two extremes and prices vary greatly ranging from $326 to $18823.

You should always analyze your dataset in the simplest way possible, build
hypotheses, and devise more specific analyses to probe those hypotheses. The
`glimpse()` and `summary()` functions are two of the simplest tools we have.

## The Pipe Operator
<!-- -------------------------------------------------- -->

Throughout this class we're going to make heavy use of the *pipe operator*
`%>%`. This handy little function will help us make our code more readable.
Whenever you see `%>%`, you can translate that into the word "then". For
instance

```{r pipe-example}
diamonds %>%
  group_by(cut) %>%
  summarize(carat_mean = mean(carat))
```

Would translate into the tiny "story"

- Take the `diamonds` dataset, *then*
- Group it by the variable `cut`, *then*
- summarize it by computing the `mean` of `carat`

*What the pipe actually does*. The pipe operator `LHS %>% RHS` takes its
left-hand side (LHS) and inserts it as the first argument to the function on
its right-hand side (RHS). So the pipe will let us take `glimpse(diamonds)` and
turn it into `diamonds %>% glimpse()`.

### __q4__ Use the pipe operator to re-write `summary(diamonds)`.

```{r q4-task}
diamonds %>% summary
```

## Reading Data
<!-- -------------------------------------------------- -->

So far we've only been looking at built-in datasets. Ultimately, we'll want to read in our own data. We'll get to the art of loading and *wrangling* data later, but for now, know that the `readr` package provides us tools to read data. Let's quickly practice loading data below.

### __q5__ Use the function `read_csv()` to load the file `"./data/tiny.csv"`.

```{r q5-task}
read_csv("./data/tiny.csv")
```

<!-- include-exit-ticket -->
# Exit Ticket
<!-- -------------------------------------------------- -->

Once you have completed this exercise, make sure to fill out the **exit ticket survey**, [linked here](https://docs.google.com/forms/d/e/1FAIpQLSeuq2LFIwWcm05e8-JU84A3irdEL7JkXhMq5Xtoalib36LFHw/viewform?usp=pp_url&entry.693978880=e-data00-basics-assignment.Rmd).
