---
title: "Gapminder"
author: "Aaron Huang"
date: 2023-2-26
output:
  github_document:
    toc: true
prerequisites:
  - e-vis03-boxplots
---

*Purpose*: Learning to do EDA well takes practice! In this challenge you'll further practice EDA by first completing a guided exploration, then by conducting your own investigation. This challenge will also give you a chance to use the wide variety of visual tools we've been learning.

<!-- include-rubric -->
# Grading Rubric
<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics define how you will be graded, both on an individual and team basis.

## Individual
<!-- ------------------------- -->

| Category | Needs Improvement | Satisfactory |
|----------|----------------|--------------|
| Effort | Some task __q__'s left unattempted | All task __q__'s attempted |
| Observed | Did not document observations, or observations incorrect | Documented correct observations based on analysis |
| Supported | Some observations not clearly supported by analysis | All observations clearly supported by analysis (table, graph, etc.) |
| Assessed | Observations include claims not supported by the data, or reflect a level of certainty not warranted by the data | Observations are appropriately qualified by the quality & relevance of the data and (in)conclusiveness of the support |
| Specified | Uses the phrase "more data are necessary" without clarification | Any statement that "more data are necessary" specifies which *specific* data are needed to answer what *specific* question |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability | Code sufficiently close to the [style guide](https://style.tidyverse.org/) |

## Due Date
<!-- ------------------------- -->

All the deliverables stated in the rubrics above are due **at midnight** before the day of the class discussion of the challenge. See the [Syllabus](https://docs.google.com/document/d/1qeP6DUS8Djq_A0HMllMqsSqX3a9dbcx1/edit?usp=sharing&ouid=110386251748498665069&rtpof=true&sd=true) for more information.

```{r setup}
library(tidyverse)
library(gapminder)

```

*Background*: [Gapminder](https://www.gapminder.org/about-gapminder/) is an
independent organization that seeks to educate people about the state of the
world. They seek to counteract the worldview constructed by a hype-driven media
cycle, and promote a "fact-based worldview" by focusing on data. The dataset
we'll study in this challenge is from Gapminder.

# Guided EDA
<!-- -------------------------------------------------- -->

First, we'll go through a round of *guided EDA*. Try to pay attention to the high-level process we're going through---after this guided round you'll be responsible for doing another cycle of EDA on your own!

### __q0__ Perform your "first checks" on the dataset. What variables are in this
dataset?

```{r q0-task}
## TASK: Do your "first checks" here!
glimpse(gapminder)
```

**Observations**:

- Write all variable names here
The variable names are country, continent, year, lifeExp, pop, and gdpPercap.

### __q1__ Determine the most and least recent years in the `gapminder` dataset.

*Hint*: Use the `pull()` function to get a vector out of a tibble. (Rather than the `$` notation of base R.)

```{r q1-task}
## TASK: Find the largest and smallest values of `year` in `gapminder`
year_max <- gapminder %>%
  select(year) %>%
  max()
year_min <- gapminder %>%
  select(year) %>%
  min()
```

Use the following test to check your work.

```{r q1-tests}
## NOTE: No need to change this
assertthat::assert_that(year_max %% 7 == 5)
assertthat::assert_that(year_max %% 3 == 0)
assertthat::assert_that(year_min %% 7 == 6)
assertthat::assert_that(year_min %% 3 == 2)

if (is_tibble(year_max)) {
  print("year_max is a tibble; try using `pull()` to get a vector")
  assertthat::assert_that(False)
}

print("Nice!")
```

### __q2__ Filter on years matching `year_min`, and make a plot of the GDP per capita against continent. Choose an appropriate `geom_` to visualize the data. What observations can you make?

You may encounter difficulties in visualizing these data; if so document your challenges and attempt to produce the most informative visual you can.

```{r q2-task}
## TASK: Create a visual of gdpPercap vs continent
gapminder %>%
  filter(year == year_min) %>%
  ggplot(aes(continent, gdpPercap)) +
  geom_col()

gapminder %>%
  filter(year == year_min) %>%
  ggplot(aes(continent, gdpPercap)) +
  geom_boxplot() +
  scale_y_log10()
```

**Observations**:

- Write your observations here
Oceania has the highest median GDP per cap, followed by Europe, Americas, Asia, and Africa. However, there are individual countries in Europe, Asia, and the Americas who have higher GDP per cap. Specifically, there is one country in Asia whose GDP per cap is significantly higher than any other country.

**Difficulties & Approaches**:

- Write your challenges and your approach to solving them
Initially I made a simple bar chart to get a better understanding of the data. Then, I created a boxplot to look at the variation in data within each continent.

### __q3__ You should have found *at least* three outliers in q2 (but possibly many more!). Identify those outliers (figure out which countries they are).

```{r q3-task}
## TASK: Identify the outliers from q2
gapminder %>%
  filter(year == year_min & continent == "Europe") %>%
  arrange(desc(gdpPercap))

gapminder %>%
  filter(year == year_min & continent == "Asia") %>%
  arrange(desc(gdpPercap))

gapminder %>%
  filter(year == year_max & continent == "Americas") %>%
  arrange(desc(gdpPercap))
```

**Observations**:

- Identify the outlier countries from q2
  - The major outlier countries are Switzerland, Kuwait, and the United States.

*Hint*: For the next task, it's helpful to know a ggplot trick we'll learn in an upcoming exercise: You can use the `data` argument inside any `geom_*` to modify the data that will be plotted *by that geom only*. For instance, you can use this trick to filter a set of points to label:

```{r layer-filter}
## NOTE: No need to edit, use ideas from this in q4 below
gapminder %>%
  filter(year == year_max) %>%
  ggplot(aes(continent, lifeExp)) +
  geom_boxplot() +
    geom_point(
    data = . %>% filter(country %in% c("United Kingdom", "Japan", "Zambia")),
    mapping = aes(color = country),
    size = 2
  )
```

### __q4__ Create a plot similar to yours from q2 studying both `year_min` and `year_max`. Find a way to highlight the outliers from q3 on your plot *in a way that lets you identify which country is which*. Compare the patterns between `year_min` and `year_max`.

*Hint*: We've learned a lot of different ways to show multiple variables; think about using different aesthetics or facets.

```{r q4-task}
## TASK: Create a visual of gdpPercap vs continent
gapminder %>%
  filter(year == year_min | year == year_max) %>%
  ggplot(aes(continent, gdpPercap)) +
  geom_boxplot() +
  geom_point(
    data = . %>% filter(country %in% c("Switzerland", "United States", "Kuwait", "Canada", "Haiti")),
    mapping = aes(color = country),
    size = 2) +
  facet_grid(~year) +
  scale_y_log10()
  
```

**Observations**:

- Write your observations here
In 1952, the top 4 outliers were the United States, Canada, Switzerland, and Kuwait. Kuwait's gdp per cap was significantly higher than any other country. The gdp per cap for the other outlier countries are closer in value. In contrast, in 2007, Kuwait's gdp per cap dropped significantly. Switzerland, Canada, and the United States has increased in gdp per cap and has caught up with Kuwait. Haiti appears to be a new outlier in the Americas in 2007 that was not present in 1952.

# Your Own EDA
<!-- -------------------------------------------------- -->

Now it's your turn! We just went through guided EDA considering the GDP per capita at two time points. You can continue looking at outliers, consider different years, repeat the exercise with `lifeExp`, consider the relationship between variables, or something else entirely.

### __q5__ Create *at least* three new figures below. With each figure, try to pose new questions about the data.

```{r q5-task1}
## TASK: Your first graph
gapminder %>%
  filter(year == year_max) %>%
  ggplot(aes(gdpPercap, lifeExp)) + 
  geom_point() + 
  facet_wrap(~continent)
```

- (Your notes and observations here)
Q: Does a larger gdp per capita lead to a longer life expectancy in 2007?

Observations:
The plots for the Americas, Asia, and Europe all look quite flat with a slight upward trend, suggesting there is not a strong correlation between life expectancy and gdp per cap. The trendline for Africa almost appears vertical suggesting there is not correlation at all. Based on these graphs, it appears there is only a very slight positive correlation between life expectancy and gdp per cap, if any.


```{r q5-task2}
## TASK: Your second graph
gapminder %>%
  filter(country == "Kuwait") %>%
  ggplot(aes(year, gdpPercap)) +
  geom_line() +
  geom_point()
```

- (Your notes and observations here)
Q: How has Kuwait's gdp per cap changed over the years?
Kuwait's GDP per cap has always been extremely high due to their plentiful oil resources, which means they were directly tied to the price of global market price of oil. When there was an oil shortage in the 1970s, the gdp per cap skyrocketed and when there was a recession in the 1980s, the gdp per cap fell significantly.

```{r q5-task3}
## TASK: Your third graph
min_max_norm <- function(x) {
    (x - min(x)) / (max(x) - min(x))
  }

gapminder_norm <- as.data.frame(lapply(gapminder[5:6], min_max_norm))
gapminder_norm$country <- gapminder$country

gapminder_norm %>%
  filter(country == "United States" | country == "United Kingdom" | 
           country == "China" | country == "Singapore") %>%
  ggplot(aes(pop, gdpPercap, color = country)) + 
  geom_point() +
  facet_wrap(~country, scales="free_x")
```

- (Your notes and observations here)
Q: Is there a correlation between the population of a country and the gdp per cap?
Looking at the normalized population and gdp per cap for a few countries, there seems to be a positive correlation between the population and gdp per cap, which might be suggestive of a causal relationship. The slope of this correlation varies greatly depending on the country, which suggests there may be other factors to better explain a country's gdp per cap.
