---
title: "SAT and College Grades"
author: "Aaron Huang"
date: 2023-4-17
output:
  github_document:
    toc: true
prerequisites:
  - e-vis00-basics
---

*Purpose*: How do we apply hypothesis testing to investigating data? In this challenge you'll practice using hypothesis testing tools to make sense of a dataset.

*Reading*:
- [Harvard Study Says SATs Should Be Optional: Here's Why](https://www.csmonitor.com/USA/USA-Update/2016/0120/Harvard-study-says-SATs-should-be-optional.-Here-s-why) (Optional); easy-to-read news article on colleges going SAT-free
- [Norm-Referenced Tests and Race-Blind Admissions](https://cshe.berkeley.edu/publications/norm-referenced-tests-and-race-blind-admissions-case-eliminating-sat-and-act-university) (Optional); technical report on relationship between the SAT/ACT and non-academic factors

*Credit*: This is based on a [case study](http://onlinestatbook.com/2/case_studies/sat.html) originally prepared by Emily Zitek, with data collected through the research of Thomas MacFarland.

```{r setup}
library(tidyverse)
library(readxl)
library(broom)
library(modelr)
library(rsample)
library(boot)
```

<!-- include-rubric -->
# Grading Rubric
<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics define how you will be graded, both on an individual and team basis.

## Individual
<!-- ------------------------- -->

| Category | Needs Improvement | Satisfactory |
|----------|----------------|--------------|
| Effort | Some task __q__'s left unattempted | All task __q__'s attempted |
| Observed | Did not document observations, or observations incorrect | Documented correct observations based on analysis |
| Supported | Some observations not clearly supported by analysis | All observations clearly supported by analysis (table, graph, etc.) |
| Assessed | Observations include claims not supported by the data, or reflect a level of certainty not warranted by the data | Observations are appropriately qualified by the quality & relevance of the data and (in)conclusiveness of the support |
| Specified | Uses the phrase "more data are necessary" without clarification | Any statement that "more data are necessary" specifies which *specific* data are needed to answer what *specific* question |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability | Code sufficiently close to the [style guide](https://style.tidyverse.org/) |

## Due Date
<!-- ------------------------- -->

All the deliverables stated in the rubrics above are due **at midnight** before the day of the class discussion of the challenge. See the [Syllabus](https://docs.google.com/document/d/1qeP6DUS8Djq_A0HMllMqsSqX3a9dbcx1/edit?usp=sharing&ouid=110386251748498665069&rtpof=true&sd=true) for more information.

*Background*: Every year about 2 million students take the Scholastic Aptitude Test (SAT). The exam is [controversial](http://www.nea.org/home/73288.htm) but [extremely consequential](https://www.csmonitor.com/2004/0518/p13s01-legn.html). There are many claims about the SAT, but we're going to look at just one: Is the SAT predictive of scholastic performance in college? It turns out this is a fairly complicated question to assess---we'll get an introduction to some of the complexities.

# Obtain the Data
<!-- -------------------------------------------------- -->

### __q1__ Visit the [SAT and College GPA](http://onlinestatbook.com/2/case_studies/sat.html) case study page, scroll to the bottom, and click the `Open Data with Excel` button. This will allow you to download an `xls` file. Save the file to your `data` folder, load the data as `df_sat`, and perform your "first checks" against these data. Answer the questions below:

```{r q1-task}
## TODO:
df_sat <- read.csv("data/sat.csv")

## TODO: Do your "first checks"
glimpse(df_sat)
```

**Observations**:

- Fill in the following "data dictionary"

| Column     | Meaning |
|------------|---------|
| `high_GPA` | High school grade point average          |
| `math_SAT` | Math SAT score                           |
| `verb_SAT` | Verbal SAT score                         |
| `comp_GPA` | Computer science grade point average     |
| `univ_GPA` | Overall university grade point average   |

- What information do we have about these students?
  - The 105 students in this study graduated from a state university with a B.S. in computer science.
- What kinds of information *do we not have* about these students?
  - We do not have information about their background, such as their ethnicity, where they went to high school at, or where they are from.
- Based on these missing variables, what possible effects could be present in the data that we would have *no way of detecting*?
  - Without knowing more information, we do not know the circumstances of their upbringing before going to college. They may have gone to a bad high school which may result in the data looking like they improved in college but in reality was just put in a better learning environment.

# Analysis with Hypothesis Testing
<!-- ----------------------------------------------------------------------- -->

We're going to use two complementary approaches to analyze the data, the first based on hypothesis testing of correlation coefficients, and the second based on fitting a regression model and interpreting the regression coefficients.

To simplify the analysis, let's look at a composite SAT score:

```{r compute-composite}
## NOTE: No need to edit this
df_composite <-
  df_sat %>%
  mutate(both_SAT = math_SAT + verb_SAT)

glimpse(df_composite)
```

## View 1: Correlations
<!-- ----------------------------------------------------------------------- -->

### __q2__ Create a *single* plot that shows `univ_GPA` against *both* `high_GPA` and `both_SAT`. Visually compare the two trends.

*Hint*: One way to do this is to first *pivot* `df_composite`.

```{r q2-task}
## TODO
df_composite %>% 
  pivot_longer(df_composite, 
               cols = c("high_GPA", "both_SAT"),
               names_to = "variable", values_to = "value") %>%
  ggplot(aes(x = univ_GPA, y = value)) +
    geom_point() +
    facet_wrap(~variable, scales = "free_y") +
    labs(x = "University GPA", y = "Value", color = "Variable")
```

**Observations**:

- What relationship do `univ_GPA` and `both_SAT` exhibit?
  - There seems to be a positive correlation between their university GPA and SAT score. The trend looks exponential, however, if you are only looking at the values with university GPA above 3, it looks like a vertical line.
- What relationship do `univ_GPA` and `high_GPA` exhibit?
  - There seems to be a positive correlation between their university GPA and highschool GPA. The trend is pretty clear and linear at the very top right corner and there is a gap of data points at about (2.75, 2.75). There appears to be some big outliers that don't fit this trend.

### Hypothesis Testing with a Correlation Coefficient
<!-- ------------------------- -->

We can use the idea of hypothesis testing with a correlation coefficient. The
idea is to set our null hypothesis to the case where there is no correlation,
and test to see if the data contradict that perspective. Formally, the null (H0)
and alternative (HA) hypotheses relating to a correlation coefficient between
two variables `X, Y` are:

$$\text{H0: } \text{Corr}[X, Y] = 0$$

$$\text{HA: } \text{Corr}[X, Y] \neq 0$$

The R function `cor.test` implements such a hypothesis test under the assumption that `X, Y` are both normally distributed. First, let's check to see if this assumption looks reasonable for our data.

### __q3__ Plot histograms for `both_SAT, high_GPA, univ_GPA`. Which---if any---of the variables look approximately normally distributed.

```{r q3-task}
bin_size <- 45

df_composite %>%
  ggplot() +
    geom_histogram(aes(x = both_SAT), bins = bin_size, color = "black", fill = "lightblue") +
    ggtitle("Histogram of Both SAT") +
    xlab("Both SAT") +
    ylab("Frequency")

df_composite %>%
  ggplot() +
    geom_histogram(aes(x = high_GPA), bins = bin_size, color = "black", fill = "lightblue") +
    ggtitle("Histogram of High School GPA") +
    xlab("High School GPA") +
    ylab("Frequency")

df_composite %>%
  ggplot() +
    geom_histogram(aes(x = univ_GPA), bins = bin_size, color = "black", fill = "lightblue") +
    ggtitle("Histogram of University GPA") +
    xlab("University GPA") +
    ylab("Frequency")
```

**Observations**:

- To what extent does `both_SAT` look like a normal distribution?
  - It does not really look like a normal distribution even when the bin size was increased. The ends looks like a normal distribution but the middle is missing the height.
- To what extent does `high_GPA` look like a normal distribution?
  - This looks more like a normal distribution compared to the previous plot, however, it is also missing height (3.0) where we would expect there to be.
- To what extent does `univ_GPA` look like a normal distribution?
  - There appears to be two peaks at about 2.3-2.4 and 3.5-3.6. With the limited data, they both appear to be in the form of a normal distribution.

Keep in mind your findings as you complete q4.

### __q4__ Use the function `cor.test()` to construct confidence intervals for `corr[high_GPA, univ_GPA` and `corr[both_SAT, univ_GPA]`. Answer the questions below.

```{r q4-task}
## TODO: Use the function cor.test() to test the correlations between
##       high_GPA and univ_GPA, as well as between
##       both_SAT and univ_GPA
cor.test(df_composite$high_GPA, df_composite$univ_GPA)
cor.test(df_composite$both_SAT, df_composite$univ_GPA)
```

**Observations**:

- Which correlations are significantly nonzero?
  - Both correlations are significantly nonzero.
- Which of `high_GPA` and `both_SAT` seems to be more strongly correlated with `univ_GPA`?
  - Their high school GPA seems to be more strongly correlated with their university GPA
- How do the results here compare with the visual you created in q2?
  - This makes sense because there is a clearer trend for high school and university gpa.
- Based on these results, what can we say about the predictive capabilities of both `high_GPA` and `both_SAT` to predict `univ_GPA`?
  - High school GPA is a better metric to look at to predict university GPA, however, the correlation for both metrics with university GPA is still pretty bad.

Finally, let's use the bootstrap to perform the same test using *different* assumptions.

### __q5__ Use the bootstrap to approximate a confidence interval for `corr[high_GPA, univ_GPA`. Compare your results---both the estimate and confidence interval---to your results from q4.

```{r q5-task}
## TODO: Use the bootstrap to compute a confidence interval for corr[high_GPA, univ_GPA]
set.seed(100)

corr <- function(data, idx)
{
  df <- data[idx, ]
  c(cor(df[, 1], df[, 5], method = 'pearson'))
}
 
bootstrap <- boot(df_composite, corr, R = 10000)
bootstrap
boot.ci(boot.out = bootstrap, type = "perc") 

```

**Observations**:

- How does your estimate from q5 compare with your estimate from q4? How does your CI from q5 compare with your CI from q4?
  - The estimate from q4 was 0.685 with a confidence interval of 0.567 and 0.775. The esimate from the bootstrap is 0.780 with a confidence interval of 0.6966 and 0.8477. The bootstrap method has a higher estimate than found in q4. The confidence interval values are different but the range are relatively the same.


*Aside*: When you use two different approximations to compute the same quantity and get similar results, that's an *encouraging sign*. Such an outcome lends a bit more credibility to the results.

## View 2: Modeling
<!-- ------------------------- -->

Correlations are useful for relating two variables at a time. To study the relationship among more variables we can instead use a fitted model. Using a model, we can also help assess whether it is *worthwhile* to measure a variable.

To begin, let's first split the data into training and validation sets.

```{r split}
## NOTE: No need to edit
set.seed(101)

df_train <-
  df_composite %>%
  rowid_to_column() %>%
  slice_sample(n = 80)

df_validate <-
  df_composite %>%
  rowid_to_column() %>%
  anti_join(
    .,
    df_train,
    by = "rowid"
  )
```

### Hypothesis Testing with a Model
<!-- ------------------------- -->

We can combine the ideas of hypothesis testing with a model. Using a model, we can express our hypotheses in terms of the model parameters. For instance, if we were interested in whether $X$ has an affect on $Y$, we might set up a model:

$$Y_i = \beta X_i + \epsilon_i$$

With the hypotheses:

$$\text{H0}: \beta = 0$$

$$\text{HA}: \beta \neq 0$$

In this case, we're testing for whether $X$ has a significant effect on $Y$. Let's apply this idea to relating the variables `univ_GPA` and `high_GPA`. Luckily R has built-in tools to construct a confidence interval on the $\beta$'s in a regression [1]; we'll simply use those tools rather than do it by hand.

### __q6__ Fit a linear model predicting `univ_GPA` with the predictor `both_SAT`. Assess the model to determine how effective a predictor `both_SAT` is for `univ_GPA`. Interpret the resulting confidence interval for the coefficient on `both_SAT`.

```{r q6-task}
## TODO: Fit a model of univ_GPA on the predictor both_SAT
fit_basic <- lm(univ_GPA ~ both_SAT, data = df_composite)

## NOTE: The following computes confidence intervals on regression coefficients
fit_basic %>%
  tidy(
    conf.int = TRUE,
    conf.level = 0.99
  )
```

**Observations**:

- What is the confidence interval on the coefficient of `both_SAT`? Is this coefficient significantly different from zero?
  - The confidence interval is 0.001984016 and 0.003491187. These numbers are small, but are not zero.
- By itself, how well does `both_SAT` predict `univ_GPA`?
  - Based on the information above, it appears both_SAT does not predict univ_GPA very well, considering how high the p-value is.

Remember from `e-model03-interp-warnings` that there are challenges with interpreting regression coefficients! Let's investigate that idea further.

### __q7__ Fit a model predicting `univ_GPA` using both `high_GPA` and `both_SAT`. Compare the prediction accuracy and hypothesis test results.

```{r q7-task}
## TODO: Fit and assess models with predictors both_SAT + high_GPA, and high_GPA alone
fit_basic <- lm(univ_GPA ~ both_SAT + high_GPA, data = df_composite)

## NOTE: The following computes confidence intervals on regression coefficients
fit_basic %>%
  tidy(
    conf.int = TRUE,
    conf.level = 0.99
  )
```

**Observations**:

- How well do these models perform, compared to the one you built in q6?
  - This model appears to perform slightly better than the previous model in 16 because the p-value is one order of magnitude smaller.
- What is the confidence interval on the coefficient of `both_SAT` when including `high_GPA` as a predictor?? Is this coefficient significantly different from zero?
  - When including high_GPA, the confidence interval range has shrunk compared to q5, especially the lower confidence interval. However, the upper confidence interval is not zero.
- How do the hypothesis test results compare with the results in q6?
  - Since the p-value is above 0.05 for both models, we cannot reject the null hypothesis. However, it is important to note the modelin q7 is better are predicting the value university GPA.

## Synthesize
<!-- ------------------------- -->

Before closing, let's synthesize a bit from the analyses above.

### __q8__ Using the results from all previous q's, answer the following questions.

**Observations**:

- Between `both_SAT` and `high_GPA`, which single variable would you choose to predict `univ_GPA`? Why?
  - I would choose high_GPA to predict univ_GPA because looking at the model from q7, there appears to be a correlation between univ_GPA and high_GPA. The estimate for both_SAT is essentially 0 while the estimate for high_GPA is non-zero. This means that the value for univ_GPA depends highly on high_GPA and not as much on both_SAT.
- Is `both_SAT` an effective predictor of `univ_GPA`? What specific pieces of evidence do you have in favor of `both_SAT` being effective? What specific pieces of evidence do you have against?
  - both_SAT is not an effective predictor of univ_GPA. This can be seen in the models built in q6 and q7. The p-value in q6 tells us there is not a big correlation between the two. The model built in q7 shows that both_SAT does not contribute significantly to value of univ_GPA as the estimate value is 0.0008 while the estimate for high_GPA is 0.541.

# End Notes
<!-- ----------------------------------------------------------------------- -->

[1] There are also assumptions underlying this kind of testing, for more information see this [Wiki article](https://en.wikipedia.org/wiki/Linear_regression#Assumptions).
