---
title: "Massachusetts Highway Stops"
author: "Aaron Huang"
date: 2023-5-5
output:
  github_document:
    toc: true
---

*Purpose*: In this last challenge we'll focus on using logistic regression to study a large, complicated dataset. Interpreting the results of a model can be challenging---both in terms of the statistics and the real-world reasoning---so we'll get some practice in this challenge.

<!-- include-rubric -->
# Grading Rubric
<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics define how you will be graded, both on an individual and team basis.

## Individual
<!-- ------------------------- -->

| Category | Needs Improvement | Satisfactory |
|----------|----------------|--------------|
| Effort | Some task __q__'s left unattempted | All task __q__'s attempted |
| Observed | Did not document observations, or observations incorrect | Documented correct observations based on analysis |
| Supported | Some observations not clearly supported by analysis | All observations clearly supported by analysis (table, graph, etc.) |
| Assessed | Observations include claims not supported by the data, or reflect a level of certainty not warranted by the data | Observations are appropriately qualified by the quality & relevance of the data and (in)conclusiveness of the support |
| Specified | Uses the phrase "more data are necessary" without clarification | Any statement that "more data are necessary" specifies which *specific* data are needed to answer what *specific* question |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability | Code sufficiently close to the [style guide](https://style.tidyverse.org/) |

## Due Date
<!-- ------------------------- -->

All the deliverables stated in the rubrics above are due **at midnight** before the day of the class discussion of the challenge. See the [Syllabus](https://docs.google.com/document/d/1qeP6DUS8Djq_A0HMllMqsSqX3a9dbcx1/edit?usp=sharing&ouid=110386251748498665069&rtpof=true&sd=true) for more information.

*Background*: We'll study data from the [Stanford Open Policing Project](https://openpolicing.stanford.edu/data/), specifically their dataset on Massachusetts State Patrol police stops.

```{r setup}
library(tidyverse)
library(broom)
```

# Setup
<!-- -------------------------------------------------- -->

### __q1__ Go to the [Stanford Open Policing Project](https://openpolicing.stanford.edu/data/) page and download the Massachusetts State Police records in `Rds` format. Move the data to your `data` folder and match the `filename` to load the data.

*Note*: An `Rds` file is an R-specific file format. The function `readRDS` will read these files.

```{r q1-task}
## TODO: Download the data, move to your data folder, and load it
filename <- "data/ma_statewide_2020_04_01.rds"
df_data <- readRDS(filename)
```

# EDA
<!-- -------------------------------------------------- -->

### __q2__ Do your "first checks" on the dataset. What are the basic facts about this dataset?

```{r q2-task}
## TODO: Determine the factor levels for subject_race and raw_Race
glimpse(df_data)
head(df_data)
summary(df_data)
```
There are a total of 24 columns and 3.4+ rows of data. The columns are row_row_number, date, location, county_name, subject_age, subject_race, subject_sex, type, arrest_made, citation_issued, warning_issued, outcome, contraband_found, contraband_drugs, contraband_weapons, contraband_alcohol, contraband_other, frisk_performed, search_conducted, search_basis, reason_for_stop, vehicle_type, vehicle_registration_state, and raw_race. There seems to be lots of NA for certain columns such as column names with contraband.

Note that we have both a `subject_race` and `race_Raw` column. There are a few possibilities as to what `race_Raw` represents:

- `race_Raw` could be the race of the police officer in the stop
- `race_Raw` could be an unprocessed version of `subject_race`

Let's try to distinguish between these two possibilities.

### __q3__ Check the set of factor levels for `subject_race` and `raw_Race`. What do you note about overlap / difference between the two sets?

```{r q3-task}
## TODO: Determine the factor levels for subject_race and raw_Race
df_data %>%
  select(subject_race) %>%
  unique()

df_data %>%
  select(raw_Race) %>%
  unique()

```

**Observations**:

- What are the unique values for `subject_race`?
  The unique values for subject_race are white, hispanic, black, asian/pacific islander, other, NA, and unknown
- What are the unique values for `raw_Race`?
  The unique values for raw_Race are white, hispanic, black, asian or pacific islander, middle eastern or east indian, american indian or alaskan native, NA, none, and A
- What is the overlap between the two sets?
  The overlap between the two are white, hispanic, black, asian/pacific islander, and NA.
- What is the difference between the two sets?
  The unique values only found in subject_race are other and unknown. The unique values only found in raw_Race are middle eastern or east indian, american Indian or Alaskan native, none, and A.

### __q4__ Check whether `subject_race` and `raw_Race` match for a large fraction of cases. Which of the two hypotheses above is most likely, based on your results?

*Note*: Just to be clear, I'm *not* asking you to do a *statistical* hypothesis test.

```{r q4-task}
## TODO: Devise your own way to test the hypothesis posed above.
df_data %>%
  select(subject_race, raw_Race) %>%
  filter(raw_Race == "A")

df_data %>%
  select(subject_race, raw_Race) %>%
  filter(raw_Race == "None - for no operator present citations only")
```

**Observations**

Between the two hypotheses:

- `race_Raw` could be the race of the police officer in the stop
- `race_Raw` could be an unprocessed version of `subject_race`

which is most plausible, based on your results?

- race_Raw could be an unprocessed version of subject_race because comparing the two columns, most of the values are the same except for capital letters in raw_Race and lowercase letters in subject_race. Middle Eastern or East Indian is converted to asian/pacific islander, American Indian or Alaskan Native and A are converted to other, and none is converted to unknown. These conversions are equivalent in value which suggests race_Raw is the unprocessed version.

## Vis
<!-- ------------------------- -->

### __q5__ Compare the *arrest rate*---the fraction of total cases in which the subject was arrested---across different factors. Create as many visuals (or tables) as you need, but make sure to check the trends across all of the `subject` variables. Answer the questions under *observations* below.

(Note: Create as many chunks and visuals as you need)

```{r q5-age}
# Calculate arrest rate by age
df_data_clean <- df_data[complete.cases(df_data$subject_age, df_data$arrest_made), ]
df_data_clean <- df_data_clean %>%
  mutate(age_group = cut(subject_age, breaks = c(0, 25, 30, 35, 40, 45, 50,Inf), 
                         labels = c("18-25", "26-30", "31-35", "36-40", "41-45", "46-50", "51+"), right = FALSE))


arrest_rate_by_age <- df_data_clean %>%
  group_by(age_group) %>%
  summarize(arrest_rate = mean(arrest_made))

arrest_rate_by_age %>%
  ggplot(aes(age_group, arrest_rate)) +
  geom_col()

```
```{r q5-sex}
# Calculate arrest rate by sex
df_data_clean <- df_data[complete.cases(df_data$subject_sex, df_data$arrest_made), ]

arrest_rate_by_sex <- df_data_clean %>%
  group_by(subject_sex) %>%
  summarize(arrest_rate = mean(arrest_made))

arrest_rate_by_sex %>%
  ggplot(aes(subject_sex, arrest_rate)) +
  geom_col()

```
```{r q5-race}
# Calculate arrest rate by race
df_data_clean <- df_data[complete.cases(df_data$subject_race, df_data$arrest_made), ]

arrest_rate_by_race <- df_data_clean %>%
  group_by(subject_race) %>%
  summarize(arrest_rate = mean(arrest_made))

arrest_rate_by_race %>%
  ggplot(aes(subject_race, arrest_rate)) +
  geom_col()

```
**Observations**:

- How does `arrest_rate` tend to vary with `subject_age`?
  - People in the age range group from 26 to 35 tend to have higher arrest rates compared to other age groups. Arrest rates jump from the 18-25 age group to the next age group. Arrest rates for age groups after 31-35 decreases as age increases.
- How does `arrest_rate` tend to vary with `subject_sex`?
  - The arrest rate for male subjects are more than double than that of female subjects.
- How does `arrest_rate` tend to vary with `subject_race`?
  - Arrest rates for Hispanics are higher than any other race group, followed by black people. Arrest rates for asian, white, and other are pretty similar. 

# Modeling
<!-- -------------------------------------------------- -->

We're going to use a model to study the relationship between `subject` factors and arrest rate, but first we need to understand a bit more about *dummy variables*

### __q6__ Run the following code and interpret the regression coefficients. Answer the the questions under *observations* below.

```{r q6-task}
## NOTE: No need to edit; inspect the estimated model terms.
fit_q6 <-
  glm(
    formula = arrest_made ~ subject_age + subject_race + subject_sex,
    data = df_data %>%
      filter(
        !is.na(arrest_made),
        subject_race %in% c("white", "black", "hispanic")
      ),
    family = "binomial"
  )

fit_q6 %>% tidy()
```

**Observations**:

- Which `subject_race` levels are included in fitting the model?
  - Hispanic, white, and black are included in the fitting the model.
- Which `subject_race` levels have terms in the model?
  - The estimated coefficients for Hispanic and white are 0.513 and -0.380, which suggests individuals who are classified as Hispanic are more likely to get arrested and individuals are are classified as white are associated with lower arrest rates.

You should find that each factor in the model has a level *missing* in its set of terms. This is because R represents factors against a *reference level*: The model treats one factor level as "default", and each factor model term represents a change from that "default" behavior. For instance, the model above treats `subject_sex==male` as the reference level, so the `subject_sexfemale` term represents the *change in probability* of arrest due to a person being female (rather than male).

The this reference level approach to coding factors is necessary for [technical reasons](https://www.andrew.cmu.edu/user/achoulde/94842/lectures/lecture10/lecture10-94842.html#why-is-one-of-the-levels-missing-in-the-regression), but it complicates interpreting the model results. For instance; if we want to compare two levels, neither of which are the reference level, we have to consider the difference in their model coefficients. But if we want to compare all levels against one "baseline" level, then we can relevel the data to facilitate this comparison.

By default `glm` uses the first factor level present as the reference level. Therefore we can use `mutate(factor = fct_relevel(factor, "desired_level"))` to set our `"desired_level"` as the reference factor.

### __q7__ Re-fit the logistic regression from q6 setting `"white"` as the reference level for `subject_race`. Interpret the the model terms and answer the questions below.

```{r q7-task}
## TODO: Re-fit the logistic regression, but set "white" as the reference
## level for subject_race
# Refit the model with "white" as the reference level for subject_race

fit_q7_refit <- glm(
  formula = arrest_made ~ subject_age + subject_race + subject_sex,
  data = df_data %>%
      filter(
        !is.na(arrest_made),
        subject_race %in% c("white", "black", "hispanic", "asian/pacific islander")
      ) %>% 
      mutate(subject_race = fct_relevel(subject_race, "white")),
  family = "binomial"
)

# Print out the updated model coefficients
fit_q7_refit %>% tidy()
```

**Observations**:

- Which `subject_race` level has the highest probability of being arrested, according to this model? Which has the lowest probability?
  - Hispanics have the highest probability of being arrested with a coefficient of 0.892 and asians have the lowest probability of being arrested with a coefficient of -0.389 according to this model.
- What could explain this difference in probabilities of arrest across race? List **multiple** possibilities.
  - Three possible reasons for the difference in probabilities of arrest across race could be due to discrimination, differential policing, or socioeconomic factors. Law enforcement officers may be more likely to arrest individuals of a certain race due to prejudice or unconscious bias. Differential policing may result in policing in areas with higher crime areas where higher levels of ethnic groups may residing, leading for higher arrests for those races. Differences in socioeconomic factors such as education, income, or employment that are correlated with race may influence the likelihood of an arrest.
- Look at the set of variables in the dataset; do any of the columns relate to a potential explanation you listed?
  - There doesn't seem to be a direct variable that relate to a potential explanation I listed above.

One way we can explain differential arrest rates is to include some measure indicating the presence of an arrestable offense. We'll do this in a particular way in the next task.

### __q8__ Re-fit the model using a factor indicating the presence of contraband in the subject's vehicle. Answer the questions under *observations* below.

```{r q8-task}
## TODO: Repeat the modeling above, but control for whether contraband was found
## during the police stop
fit_q8 <- glm(
  formula = arrest_made ~ subject_age + subject_race + subject_sex + contraband_found,
  data = df_data %>%
      filter(
        !is.na(arrest_made),
        subject_race %in% c("white", "black", "hispanic", "asian/pacific islander")
      ) %>% 
      mutate(subject_race = fct_relevel(subject_race, "white")),
  family = "binomial"
)

# Print out the updated model coefficients
fit_q8 %>% tidy()
```

**Observations**:

- How does controlling for found contraband affect the `subject_race` terms in the model?
  - When controlling found contraband, the subject_race coefficients decreased for black and hispanic races. However, the fitted model still suggests individuals who are Hispanic have a higher probability of getting arrested. On base value, the model suggests that being black have a slightly lower chance of being arrested compared to a white individual, but, it is important to note the relatively high standard error for the estimated coefficient for raceblack. Computing a confidence interval results in the lower and upper values to be -0.122 and 0.0168 with a 99% confidence. The coefficient value for raceblack certainty lowered when controlling for contraband, however, we cannot definitively say if individuals who are black are more or less likely to be arrested compared to white individuals.
- What does the *finding of contraband* tell us about the stop? What does it *not* tell us about the stop?
  - The finding of contraband suggests the driver may be doing something illegal which may lead to an increased rate of arrest. However, it does not necessarily mean the initial stop was valid or due to racial profiling or bias.

### __q9__ Go deeper: Pose at least one more question about the data and fit at least one more model in support of answering that question.

```{r q9-task}
## TODO: Repeat the modeling above, but control for whether contraband was found
## during the police stop
fit_q9 <- glm(
  formula = arrest_made ~ contraband_drugs + contraband_alcohol + contraband_weapons + contraband_other,
  data = df_data %>%
      filter(
        !is.na(arrest_made),
        subject_race %in% c("white", "black", "hispanic", "asian/pacific islander")
      ) %>% 
      mutate(subject_race = fct_relevel(subject_race, "white")),
  family = "binomial"
)

# Print out the updated model coefficients
fit_q9 %>% tidy()
```
**Question**:
What kind of contraband are individuals more likely to get arrested for?

**Observations**:
The fitted model suggests stops where contraband other than drugs, alcohol, or weapons were found is associated with a highest rate of arrest with a coefficient of 0.617. Closely behind is alcohol with a coefficient of 0.610. Although the model suggests a higher rate of arrest when drugs or weapons were found in the vehicle (0.349 and 0.227 respectively), alcohol and other contraband (potentially than finding drugs or weapons) have a much higher rate of arrest.


## Further Reading
<!-- -------------------------------------------------- -->

- Stanford Open Policing Project [findings](https://openpolicing.stanford.edu/findings/).
